---
title: "ERA5 Climate data for Malawi"
author: "Victor & Doreen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document converts ERA5 NetCDF's into time series for Malawi.

We should create one time series for each District for further analysis.

```{r}

library(dplyr)
library(lubridate)
library(tidyr)
library(parallel)
library(ncdf4)
library(MMWRweek)
library(geodata)
library(terra)
```

We can use the code bellow to compute monthly temperature and rainfall for each district.

```{r}

#district level data
# Load Malawi districts (level 1)
mwi_districts <- geodata::gadm(country = "MWI", level = 1, path = "../data/")

# Check the column names
names(mwi_districts)

# Usually NAME_1 contains district names
plot(mwi_districts, main = "Malawi Districts (Level 1)")




```
# Loading the NetCDF datasets that we downloaded in Python.

```{r}
temp <- rast('ERA5_temperature_malawi_2015_2024.nc')  # Temperature
precip <- rast('ERA5_precipitation_malawi_2015_2024.nc')#Precipitation
wind <- rast('ERA5_wind_malawi_2015_2024.nc')
```

# Function to compute district level monthly mean

```{r}
# Function to extract district monthly means
ERA5_district_monthly <- function(nc_rast, districts, var_name = "variable") {
  
  results_list <- list()
  
  for (i in 1:nrow(districts)) {
    district_name <- districts$NAME_1[i]
    district_poly <- districts[i, ]
    
    masked_raster <- terra::mask(nc_rast, district_poly)
    
    dates <- as.Date(time(nc_rast))
    monthly_mean <- terra::global(masked_raster, fun = mean, na.rm = TRUE)
    
    df <- data.frame(
      district = district_name,
      date     = dates,
      value    = as.numeric(monthly_mean[[1]])
    )
    
    results_list[[i]] <- df
  }
  
  final_df <- do.call(rbind, results_list)
  colnames(final_df)[3] <- var_name
  return(final_df)
}


```

#computing district level monthly temperature

```{r}

district_monthly_temp   <- ERA5_district_monthly(temp, mwi_districts, var_name = "temperature")
district_monthly_precip <- ERA5_district_monthly(precip, mwi_districts, var_name = "precipitation")
district_monthly_wind    <- ERA5_district_monthly(wind, mwi_districts, var_name = "wind")


```
```{r}
head(district_monthly_temp)
head(district_monthly_precip)
head(district_monthly_wind)

```
```{r}
combined_df <- district_monthly_temp %>%
  left_join(district_monthly_precip, by = c("district", "date")) %>%
  left_join(district_monthly_wind, by = c("district", "date"))

head(combined_df)

```

```{r}
write.csv(combined_df, "Malawi_district_climate_2015_2024.csv", row.names = FALSE)

```


